### 2014-10-02 tex/proofs.tex
Renamed location mutual exclusion to detemrnism

There are two slight complications in the proof: first, the field
access rule relies on the $S$-relation that operates on a different
state than then $\varsigma$-relation; and second, constructing the new
homomorphism in the equals-reference rule relies on the incoming heap
being location mutual exclusive. The $S$-relation effectively produces
an intermediary state that is the state where uninitialized references
are initialized before the field access actually takes place. In
essence, a state on the left of the $\varsigma$-relation that is a
field access, undergoes a transition in which its heap is changed to
initialize fields. Once the fields are initialized, then the actual
field access takes place. The state with the heap that has the
initialized fields is the intermediary state between the state on the
left of the $\varsigma$-relation and the state on the right of the
$\varsigma$-relation. The proof reasons separately about this
intermediary state to prove that it too is exact.

The equals-reference proof must show that any given reference in the
heap is not able to point to two distinct locations at the same
time. If the incoming heap is able to point to two valid locations at
a given reference at the same time, then the help is
non-deterministic, and it is not possible to construct a valid
homomorphism from the existing homomorphism: which location should be
used in the map? As such, the proof first establishes that
the $\varsigma$-relation preserves location mutual exclusive in the
incoming heap. Once that is shown, the exactness of the equals
reference rule is given.

@@

\begin{lemma}[$\rightarrow^*_S$ preserves location mutual exclusion]
\label{lem:S-mutex}
Given a location mutual exclusive heap, $(\cfgnt{L}_0\ \cfgnt{R}_0)$, from a
summary state with a reference $\cfgnt{r}$ and field $\cfgnt{f}$, the
new heap, $(\cfgnt{L}^\prime\ \cfgnt{R}^\prime)$, from the summary
initialization machine, $(\cfgnt{L}_0\ \cfgnt{R}_0\ \cfgnt{r}\ \cfgnt{f})
\rightarrow^*_S
(\cfgnt{L}^\prime\ \cfgnt{R}^\prime\ \cfgnt{r}\ \cfgnt{f})$, after any number of machine steps, is also
location mutual exclusive.
\end{lemma}
\begin{proof}
Induction over the number of machine steps on the summary initialization machine in \figref{fig:symInit}.

\noindent\textbf{Base Case}. The machine takes one step: $(\cfgnt{L}_0\ \cfgnt{R}_0\ \cfgnt{r}\ \cfgnt{f})
\rightarrow_S
(\cfgnt{L}_1\ \cfgnt{R}_1\ \cfgnt{r}\ \cfgnt{f})$. Let $\Lambda = \mathbb{UN}(\cfgnt{L}_0, \cfgnt{R}_0, \cfgnt{r},
\cfgnt{f})$ be the set of uninitialized locations. If $\Lambda = \emptyset$, then the \texttt{Summarize (end)}
rule is active and $(\cfgnt{L}_1\ \cfgnt{R}_1) = (\cfgnt{L}_0\ \cfgnt{R}_0)$, which is location mutual exclusive by the initial conditions in the lemma.

If $\Lambda \neq \emptyset$, then the \texttt{Summarize} rule is
active, and each new constraint location pair must be considered
individually. These pairs are partitioned into the sets
$\theta_\mathit{null}$, $\theta_\mathit{new}$,
$\theta_\mathit{alias}$, and $\theta_\mathit{orig}$ by the rule. 
\begin{itemize}
\item Any member of $\theta_\mathit{null}$, $\theta_\mathit{new}$, and
  $\theta_\mathit{alias}$ has a constraint of the form $\phi_x \wedge
  \ldots$ while any member of $\theta_\mathrm{orig}$ has a constraint
  of the form $\neg \phi_x \wedge \ldots$. As every
  $(\phi_\mathit{orig}\ \cfgnt{l}_\mathit{orig}) \in
  \cfgnt{L}_n(\cfgnt{R}_n(\cfgnt{l}_x,\cfgnt{f}))$ is location mutual exclusive
  by the initial conditions on $(\cfgnt{L}_0\ \cfgnt{R}_0)$ in the lemma statement, members of $\theta_\mathit{orig}$ are location
  mutual exclusive with each other and with members in $\theta_\mathit{null}$,
  $\theta_\mathit{new}$, and $\theta_\mathit{alias}$.
\item Any member of $\theta_\mathit{null}$ has the form $\ldots \wedge
  \cfgnt{r}_f = \cfgnt{r}_\mathit{null}$ while any member of
  $\theta_\mathit{new}$ and $\theta_\mathit{alias}$ has the form
  $\ldots \wedge \cfgnt{r}_f \neq \cfgnt{r}_\mathit{null} \wedge
  \ldots$. As $\theta_\mathit{null}$ is a singleton set, its location is
  location mutual exclusive from those in $\theta_\mathit{new}$ and
  $\theta_\mathit{alias}$.
\item Any member of $\theta_\mathit{new}$ has a constraint of the form
  $\ldots \wedge ( \wedge_{(
  \cfgnt{r}_a,\ \phi_a,\ l_a) \in \rho}
  \cfgnt{r}_f \ne \cfgnt{r}_a) )$ to assert it does not
  alias anything while any member of $\theta_\mathit{alias}$ has
  the form $\ldots \wedge \cfgnt{r}_f = \cfgnt{r}_a \wedge \ldots$ for
  some alias $\cfgnt{r}_a$ with both partitions reasoning over the same set of aliases $\rho$.  As $\theta_\mathit{new}$ is a singleton set, its
  location is location mutual exclusive from any member of
  $\theta_\mathit{alias}$.
\item Any member of $\theta_\mathit{alias}$ has the form
\[\ldots \wedge \cfgnt{r}_f = \cfgnt{r}_a \wedge ( \wedge_{( \cfgnt{r}^{\prime}_a\ \phi^{\prime}_a\ l^{\prime}_a)  \in \rho\ ( \cfgnt{r}^\prime_a \neq \cfgnt{r}_a) } \cfgnt{r}_f \neq \cfgnt{r}^{\prime}_a)\]
Each is mutually exclusive over the particular alias it asserts.
\end{itemize}
Every member in $\theta = \theta_\mathit{null} \cup
\theta_\mathit{new} \cup \theta_\mathit{alias} \cup
\theta_\mathit{orig}$ is thus location mutual exclusive making $(\cfgnt{L}_1\ \cfgnt{R}_1\ \cfgnt{r}\ \cfgnt{f})$ location mutual exclusive as well.

@@

\begin{lemma}[$\rightarrow_\mathit{FA}$ preserves location mutual exclusion]
\label{lem:FA-mutex}
Given a summary state, $s_\varsigma$, with a location mutual exclusive
heap, $(\cfgnt{L}_\varsigma\ \cfgnt{R}_\varsigma) =
\mathrm{heap}(s_\varsigma)$, the new heap,
$(\cfgnt{L}_\varsigma^\prime\ \cfgnt{R}_\varsigma^\prime) =
\mathrm{heap}(s_\varsigma^\prime)$, in any related state on the field
access rule, $s_\varsigma \rightarrow_\mathit{FA} s_\varsigma^\prime$, is also location mutual exclusive.
\end{lemma}
\begin{proof}
Proof by definition of $\rightarrow_\mathit{FA}$ in \figref{fig:sym}.

\lemref{lem:S-mutex} establishes that the heap coming out of the
$S$-relation is location mutual exclusive, so it is only needed to
show that location mutual exclusion is preserved by the call to the value function
$\mathbb{VS}$ in the rule.

\defref{def:VS} is the value function definition. From the definition,
each constraint in each member of the value set has the form
$(\phi\wedge \phi^\prime\ l)$ where $\phi$ and $\phi^\prime$ are the
access constraints. The first constraint $\phi$ represents the
condition under which the location is the location referenced by
$r$. The second constraint, predicated on having resolved the actual
location $\cfgnt{l}$ (expressed in $\phi$), is the condition under
which the location $\cfgnt{l}^\prime$ is the actual object for the
field as the reference for that field, $\cfgnt{r}^\prime =
\cfgnt{R}(\cfgnt{l},\cfgnt{f})$, may point to several different
locations.

Choose any two distinct members of the value set, $(\phi_\alpha \wedge
\phi_\alpha^\prime\ \ l_\alpha)$ and $(\phi_\beta \wedge
\phi_\beta^\prime\ \ l_\beta)$. 
\begin{itemize}
\item If $\phi_\alpha = \phi_\beta$, then $\phi_\alpha^\prime \wedge
  \phi_\beta^\prime = \cfgt{false}$ because $\cfgnt{l}_\alpha$ and
  $\cfgnt{l}_\beta$ come from the same field reference, and since the
  heap, $(\cfgnt{L}\ \cfgnt{R})$, is location mutual exclusive, locations
  under a common reference must be mutual exclusive.
\item If $\phi_\alpha \ne \phi_\beta$, then $\phi_\alpha \wedge \phi_\beta = \cfgt{false}$ because $\phi_\alpha$ comes from a location connected to $r$ that is different that the location connected to $r$ for $\phi_\beta$, and since the
  heap, $(\cfgnt{L}\ \cfgnt{R})$, is location mutual exclusive, locations
  under a common reference must be mutual exclusive.
\end{itemize}
The only change to the heap after the $S$-relation is applied is the addition of the new reference $\cfgnt{r}^\prime$ mapping to the value set from the value function. Since that set is location mutual exclusive and connect to $r^\prime$, the new heap $(\cfgnt{L}_\varsigma^\prime\ \cfgnt{R}_\varsigma^\prime) =
\mathrm{heap}(s_\varsigma^\prime)$ is also location mutual exclusive.
\end{proof}

>>>>

There are two slight complications in the proof: first, the field
access rule relies on the $S$-relation that operates on a different
state than then $\varsigma$-relation; and second, constructing the new
homomorphism in the equals-reference rule relies on the incoming heap
being deterministic. The $S$-relation effectively produces
an intermediary state that is the state where uninitialized references
are initialized before the field access actually takes place. In
essence, a state on the left of the $\varsigma$-relation that is a
field access, undergoes a transition in which its heap is changed to
initialize fields. Once the fields are initialized, then the actual
field access takes place. The state with the heap that has the
initialized fields is the intermediary state between the state on the
left of the $\varsigma$-relation and the state on the right of the
$\varsigma$-relation. The proof reasons separately about this
intermediary state to prove that it too is exact.

The equals-reference proof must show that any given reference in the
heap is not able to point to two distinct locations at the same
time. If the incoming heap is able to point to two valid locations at
a given reference at the same time, then the help is
non-deterministic, and it is not possible to construct a valid
homomorphism from the existing homomorphism: which location should be
used in the map? As such, the proof first establishes that the
$\varsigma$-relation preserves determinism when the incoming heap is
deterministic. Once that is shown, the exactness of the equals
reference rule is given.

@@

\begin{lemma}[$\rightarrow^*_S$ preserves heap determinism]
\label{lem:S-mutex}
Given a deterministic heap, $(\cfgnt{L}_0\ \cfgnt{R}_0)$, from a
summary state with a reference $\cfgnt{r}$ and field $\cfgnt{f}$, the
new heap, $(\cfgnt{L}^\prime\ \cfgnt{R}^\prime)$, from the summary
initialization machine, $(\cfgnt{L}_0\ \cfgnt{R}_0\ \cfgnt{r}\ \cfgnt{f})
\rightarrow^*_S
(\cfgnt{L}^\prime\ \cfgnt{R}^\prime\ \cfgnt{r}\ \cfgnt{f})$, after any number of machine steps, is also deterministic.
\end{lemma}
\begin{proof}
Induction over the number of machine steps on the summary initialization machine in \figref{fig:symInit}.

\noindent\textbf{Base Case}. The machine takes one step: $(\cfgnt{L}_0\ \cfgnt{R}_0\ \cfgnt{r}\ \cfgnt{f})
\rightarrow_S
(\cfgnt{L}_1\ \cfgnt{R}_1\ \cfgnt{r}\ \cfgnt{f})$. Let $\Lambda = \mathbb{UN}(\cfgnt{L}_0, \cfgnt{R}_0, \cfgnt{r},
\cfgnt{f})$ be the set of uninitialized locations. If $\Lambda = \emptyset$, then the \texttt{Summarize (end)}
rule is active and $(\cfgnt{L}_1\ \cfgnt{R}_1) = (\cfgnt{L}_0\ \cfgnt{R}_0)$, which is deterministic by the initial conditions in the lemma.

If $\Lambda \neq \emptyset$, then the \texttt{Summarize} rule is
active, and each new constraint location pair must be considered
individually. These pairs are partitioned into the sets
$\theta_\mathit{null}$, $\theta_\mathit{new}$,
$\theta_\mathit{alias}$, and $\theta_\mathit{orig}$ by the rule. 
\begin{itemize}
\item The original heap is
  deterministic by definition, so any constraint in any member of the
  set must have some term such that
  \[\begin{array}{l}
     \forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \theta_\mathit{orig}\ \\
     \ \ \ \ ((\cfgnt{l} \neq \cfgnt{l}^\prime \vee \phi \neq \phi^\prime) \Rightarrow (\phi \wedge \phi^\prime = \cfgt{false}))
     \end{array}
  \]
  Further, any member of $\theta_\mathrm{orig}$ has a constraint
  of the form $\phi = \neg \phi_x \wedge \ldots$ while any member of $\theta_\mathit{null}$, $\theta_\mathit{new}$, and
  $\theta_\mathit{alias}$ has a constraint of the form $\phi^\prime = \phi_x \wedge
  \ldots$; thus
  \[\begin{array}{l}
     \forall (\phi\ \cfgnt{l}) \in \theta_\mathit{orig}\ (\forall (\phi^\prime\ \cfgnt{l}^\prime) \in \theta_\mathit{null} \cup \theta_\mathit{new} \cup \theta_\mathit{alias}\ (\\
     \ \ \ \ \phi \wedge \phi^\prime = \cfgt{false}))
    \end{array}
  \]
\item The only member of $\theta_\mathit{null}= \{(\phi\ \cfgnt{l}_\mathit{null})\}$ has the form $\phi = \ldots \wedge
  \cfgnt{r}_f = \cfgnt{r}_\mathit{null}$ while any member of
  $\theta_\mathit{new}$ and $\theta_\mathit{alias}$ has the form
  $\phi^\prime = \ldots \wedge \cfgnt{r}_f \neq \cfgnt{r}_\mathit{null} \wedge
  \ldots$; thus
  \[\begin{array}{l}
     \forall (\phi^\prime\ \cfgnt{l}^\prime) \in \theta_\mathit{new} \cup \theta_\mathit{alias}\ (
     \phi \wedge \phi^\prime = \cfgt{false})
    \end{array}
  \]
\item The only member of $\theta_\mathit{new} =  \{(\phi\ \cfgnt{l}_f)\}$ has a constraint of the form
  $\phi = \ldots \wedge ( \wedge_{( \cfgnt{r}_a,\ \phi_a,\ l_a) \in \rho}
  \cfgnt{r}_f \ne \cfgnt{r}_a) )$ to assert it does not alias
  anything, while any member of $\theta_\mathit{alias}$ has the form
  $\phi^\prime = \ldots \wedge \cfgnt{r}_f = \cfgnt{r}_a \wedge \ldots$ to assert it aliases some
  $\cfgnt{r}_a$ with both partitions reasoning over the same set
  of aliases $\rho$; thus
  \[\begin{array}{l}
     \forall (\phi^\prime\ \cfgnt{l}^\prime) \in \theta_\mathit{alias}\ (
     \phi \wedge \phi^\prime = \cfgt{false})
    \end{array}
  \]
\item Any member of $\theta_\mathit{alias}$ has the form
\[\ldots \wedge \cfgnt{r}_f = \cfgnt{r}_a \wedge ( \wedge_{( \cfgnt{r}^{\prime}_a\ \phi^{\prime}_a\ l^{\prime}_a)  \in \rho\ ( \cfgnt{r}^\prime_a \neq \cfgnt{r}_a) } \cfgnt{r}_f \neq \cfgnt{r}^{\prime}_a)\]
And thus, 
  \[\begin{array}{l}
     \forall (\phi\ \cfgnt{l}),(\phi^\prime\ \cfgnt{l}^\prime) \in \theta_\mathit{alias}\\
     \ \ \ \ ((\cfgnt{l} \neq \cfgnt{l}^\prime \vee \phi \neq \phi^\prime) \Rightarrow (\phi \wedge \phi^\prime = \cfgt{false}))
     \end{array}
  \]
\end{itemize}
As $\theta$ is mapped to a single reference $r_f$ in an already deterministic heap, the
resulting heap $(\cfgnt{L}_1\ \cfgnt{R}_1)$ is likewise deterministic.

@@

\begin{lemma}[$\rightarrow_\mathit{FA}$ preserves heap determinism]
\label{lem:FA-mutex}
Given a state, $s_\varsigma$, with a deterministic heap,
$(\cfgnt{L}_\varsigma\ \cfgnt{R}_\varsigma) =
\mathrm{heap}(s_\varsigma)$, the new heap,
$(\cfgnt{L}_\varsigma^\prime\ \cfgnt{R}_\varsigma^\prime) =
\mathrm{heap}(s_\varsigma^\prime)$, in any state related by the field
access rule, $s_\varsigma \rightarrow_\mathit{FA} s_\varsigma^\prime$,
is also deterministic.
\end{lemma}
\begin{proof}
Proof by definition of $\rightarrow_\mathit{FA}$ in \figref{fig:sym}.

\lemref{lem:S-mutex} establishes that the heap coming out of the
$S$-relation is deterministic, so it is only needed to show that
determinism is preserved by the call to the value function,
$\mathbb{VS}$, in the rule.

Recall from \defref{def:VS} that each constraint in each member of the value set has the form
$(\phi\wedge \phi^\prime\ l)$. Choose any two distinct members of the value set, $(\phi_\alpha \wedge
\phi_\alpha^\prime\ l_\alpha^\prime)$ and $(\phi_\beta \wedge
\phi_\beta^\prime\ l_\beta^\prime)$.
\begin{itemize}
\item If $\phi_\alpha = \phi_\beta$, then by \defref{lem:S-mutex} 
\[\begin{array}{l}
\exists (\phi_\alpha\ \cfgnt{l}) \in \cfgnt{L}(\cfgnt{r})\ (\exists \cfgnt{r}^\prime \in \cfgnt{R}(\cfgnt{l},\cfgnt{f})\ (\\
\ \ \ \ (\phi_\alpha^\prime\ l_\alpha^\prime) \in \cfgnt{L}(\cfgnt{r}^\prime) \wedge (\phi_\beta^\prime\ l_\beta^\prime) \in \cfgnt{L}(\cfgnt{r}^\prime)))
\end{array}
\]
As $(\phi_\alpha^\prime\ \cfgnt{l}_\alpha^\prime)$ and $(\phi_\beta^\prime\ \cfgnt{l}_\beta^\prime)$ are distinct and connected to the same reference $\cfgnt{r}^\prime$ in a deterministic heap, $\phi_\alpha^\prime \wedge  \phi_\beta^\prime = \cfgt{false}$ by definition.
\item If $\phi_\alpha \ne \phi_\beta$, then by \defref{lem:S-mutex} 
\[\begin{array}{l}
\exists \cfgnt{l}\ ((\phi_\alpha\ \cfgnt{l}) \in \cfgnt{L}(\cfgnt{r}) \wedge \exists \cfgnt{l}^\prime \neq \cfgnt{l} \ ((\phi_\beta\ \cfgnt{l}^\prime) \in \cfgnt{L}(\cfgnt{r})))
\end{array}
\]
As $(\phi_\alpha\ \cfgnt{l})$ and $(\phi_\beta\ \cfgnt{l}^\prime)$ are distinct and connected to the same reference $\cfgnt{r}$ in a deterministic heap, $\phi_\alpha \wedge  \phi_\beta = \cfgt{false}$ by definition.
\end{itemize}
The only change to the heap after the $S$-relation is the
addition of the new reference $\cfgnt{r}^\prime$ to point to the value
set. As the value set meets the conditions for determinism, the new heap $(\cfgnt{L}_\varsigma^\prime\ \cfgnt{R}_\varsigma^\prime) =
\mathrm{heap}(s_\varsigma^\prime)$ is also deterministic.
\end{proof}
